---
title: "Data Wrangling BRFSS Data"
author: "Jorge R. Soldevila Irizarry"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Disable scientific notation globally
options(scipen = 999)
```

# Library Setup

```{r Library setup, warning=FALSE,message=FALSE}

# Install pacman if not already installed on the system. This will allow us to install desired libraries. 
if (!require("pacman")) install.packages("pacman")

# Load libraries using p_load(), which installs missing packages and loads
#libraries.
pacman::p_load(tidyverse,     # For data manipulation
               janitor,       # For cleaning column names
               survey,        # For analyzing complex survey designs
               broom,         # For tidying statistical model outputs into 
                              # data frames
               haven,         # For reading SAS Transport files, like BRFSS data
               cdlTools,      # For converting FIPS codes to state abbreviations
               expss,         # For adding variable and value labels to data frames
               openxlsx,      # For exporting data frames to Excel files
               glue,           # For easy string interpolation
               here           #For dynamic file path
               )
```

# State Parameter Setup

Define states of interest. Our analysis focuses on Hispanic population
in states where Puerto Ricans are most present: Florida and New York. As
well as analyzing Hispanics in Puerto Rico.

```{r State setup, message=FALSE, warning=FALSE}

#Define a vector containing numeric FIPS codes of state of interest.
#To find FIPS code for desired state visit: 
#https://transition.fcc.gov/oet/info/maps/census/fips/fips.txt

#We are interested in three states and state equivalent: Puerto Rico, Florida, 
#and New York
target_fips <- c(12, #Florida
                 36, #New York
                 72) #Puerto Rico


```

# Data Loading and Wrangling

## Path configuration

```{r Data and Output path, warning=FALSE, message=FALSE}
# Use here() to point to the data and output folders relative to the project root
data_path <- here("data")
output_path <- here("data") # Saving the processed RDS back to the data folder
```

## Read in Data

Here we call our data from our local folder and select desired
variables. Additionally we filter our data sets for our desired states
using the state vector created in the previous section. The objects
containing 2016 and 2023 are combined for use in the analysis.

```{r Data prep, warning=FALSE, message=FALSE}

# --- 2016 Data Processing ---

# Read and subset 2016
df_2016 <- read_xpt(file.path(data_path, "LLCP2016.XPT")) %>% 
  select(`_STATE`,`_PSU`,`_STSTR`,`_LLCPWT`,`_HISPANC`,`_MRACE1`, GENHLTH,
         `_PHYS14D`,ADDEPEV2, `_LTASTH1`, DIABETE3, `_MICHD`, CHCSCNCR, CHCOCNCR,
         HLTHPLN1) %>% 
  filter(`_STATE` %in% target_fips) %>%
  mutate(year = factor(2016)) %>%
  clean_names()

# --- 2023 Data Processing ---

# Read, subset, and standardize 2023 column names to match 2016.
df_2023 <- read_xpt(file.path(data_path, "LLCP2023.XPT")) %>% 
  select(`_STATE`,`_PSU`,`_STSTR`,`_LLCPWT`,`_HISPANC`,`_MRACE1`, GENHLTH,
         `_PHYS14D`,ADDEPEV3, `_LTASTH1`, DIABETE4, `_MICHD`, CHCSCNC1, CHCOCNC1,
         `_HLTHPL1`) %>% 
  filter(`_STATE` %in% target_fips) %>%
  rename(ADDEPEV2 = ADDEPEV3, 
         DIABETE3 = DIABETE4,
         CHCSCNCR = CHCSCNC1,
         CHCOCNCR = CHCOCNC1,
         HLTHPLN1 = `_HLTHPL1`) %>% 
  mutate(year = factor(2023)) %>%
  clean_names()

# --- Combine Data ---

# Stack the 2016 and 2023 datasets on top of each other into one large dataframe
df_combined <- bind_rows(df_2016, df_2023)

# Convert state FIPS to Abbreviation
df_combined$state <- fips(df_combined$state, to = 'Abbreviation')
```

# Create new variables

We created simplified versions of certain variables to facilitate
analysis. Additionally we create a Race and Ethnicity variable upon
which our analysis will be based on.

```{r New vars, message=FALSE, warning=FALSE}

    # Group simplified variables for General Health and Diabetes responses
df_combined <- df_combined %>%
  mutate(
    genhlth2 = case_when(genhlth %in% 1:3 ~ 1,
                         genhlth %in% 4:5 ~ 2,
                         TRUE ~ genhlth),
    diabetes = case_when(diabete3 %in% 1:2 ~ 1,
                         diabete3 == 3 ~ 2,
                         TRUE ~ diabete3),
    
    # Create a new variable that combines Race and Hispanic variables to create a 
    #new race and ethnicity variable. 
    racexhisp = case_when(
      hispanc == 1 & mrace1 %in% 1:7 ~ mrace1,
      hispanc == 1 & mrace1 == 77 ~ 8,
      hispanc == 1 & mrace1 == 99 ~ 9,
      hispanc == 2 & mrace1 %in% 1:7 ~ mrace1 + 9,
      hispanc == 2 & mrace1 == 77 ~ 17,
      hispanc == 2 & mrace1 == 99 ~ 18,
      TRUE ~ NA_real_
    )
  )
```

# Adding labels

Labels are added to our variables to ease interpretation.

```{r Labelling, message=FALSE, warning=FALSE}

#Add labels to numeric codes to faciliate analysis when results are exported.
df_combined <- apply_labels(df_combined,
  racexhisp = "Hispanic by Race",
  racexhisp = c("Hispanic White"=1, 
                "Hispanic Black"=2, 
                "Hispanic AIAN"=3, 
                "Hispanic Asian"=4, 
                "Hispanic NHPI"=5, 
                "Hispanic Other"=6, 
                "Hispanic Multirracial"=7, 
                "Hispanic Not Sure"=8, 
                "Hispanic Refused"=9, 
                "Non Hispanic White"=10, 
                "Non Hispanic Black"=11,
                "Non Hispanic AIAN"=12, 
                "Non Hispanic Asian"=13, 
                "Non Hispanic NHPI"=14, 
                "Non Hispanic Other"=15, 
                "Non Hispanic Multirracial"=16, 
                "Non Hispanic Not Sure"=17, 
                "Non Hispanic Refused"=18),
  genhlth = c("Excellent"=1, 
              "Very Good"=2, 
              "Good"=3, 
              "Fair"=4, 
              "Poor"=5, 
              "Dont know"=7, 
              "Refused"=9),
  genhlth2 = c("Good"=1, 
               "Fair/Poor"=2, 
               "Dont know"=7, 
               "Refused"=9),
  phys14d = c("Zero Days"=1, 
              "1 to 13 Days"=2, 
              "14+ Days"=3, 
              "Dont know"=9),
  addepev2 = c("Yes"=1, 
               "No"=2, 
               "Dont know"=7, 
               "Refused"=9),
  ltasth1 = c("No"=1, 
              "Yes"=2, 
              "Dont know"=7),
  diabete3 = c("Yes"=1, 
               "Yes,during pregnancy"=2, 
               "No"=3, "Pre-diabetes"=4, 
               "Dont know"=7, 
               "Refused"=9),
  diabetes = c("Yes"=1, 
               "No"=2, 
               "Dont know"=7, 
               "Refused"=9),
  michd = c("Yes"=1, 
            "No"=2),
  chcscncr = c("Yes"=1, 
               "No"=2, 
               "Dont know"=7, 
               "Refused"=9),
  chcocncr = c("Yes"=1, 
               "No"=2, 
               "Dont know"=7, 
               "Refused"=9),
  hlthpln1 = c("Yes"=1, 
               "No"=2, 
               "Dont know"=7, 
               "Refused"=9)
)

```

# Convert targeted variables to factors

```{r Convert vars to factor, message=FALSE, warning=FALSE}
# Create a list of all the variables we want to treat as categorical factors

factor_vars <- c("racexhisp", "genhlth", "genhlth2", "phys14d",
"addepev2", "ltasth1", "diabete3", "diabetes", "michd", "chcscncr",
"chcocncr", "hlthpln1")

#Convert to factor 
df_combined[factor_vars] <- lapply(df_combined[factor_vars], as.factor)

```

# Save processed data

```{r Save data, message=FALSE, warning=FALSE}
saveRDS(df_combined, file.path(output_path, "df_combined_cleaned.rds"))
```
